---
title: "Working with miniTCGA data"
author: "Palle Villesen"
date: 'Last update: `r Sys.time()`'
output: 
  html_document:
    theme: united
    code_folding: show
    toc: true
editor_options: 
  chunk_output_type: console
--- 

```{r}

library(tidyverse)

df <- read_rds(file = "prediction/TCGA_cancer_classification/miniTCGA.3349x4006.rds") 

```


```{r}



df2 <- read_rds(file = "prediction/TCGA_cancer_classification/minitcga_cancer_classification.The_Brogrammers.rds")

df3 <- read_rds(file = "prediction/ALS_data_regression/als_progression.The_Brogrammers.rds")
```

# Quick view of all data

```{r}

df[1:10,1:10]

```

# Split into training and prediction set

```{r}

set.seed(0)

data_train    <- df %>% filter(!is.na(response))
dim(data_train)

data_predict  <- df %>% filter(is.na(response))                              
dim(data_predict)

```

# Split into train and test fold

```{r}

# Always make the same split
set.seed(0)

# We use 80% for training and 20% for evaluation
trainfold <- data_train %>% sample_frac(size=0.80)
testfold  <- setdiff(data_train, trainfold)

# We fit our model (simple logistic regression on pc1 and pc2 with interaction)
fit <-  glm(response ~ pc2 * pc3, data = trainfold, family=binomial(link='logit'))

# We predict on the test fold
predicted <- predict(fit, newdata = testfold, type = "response")
predicted <- round(predicted)+1 # Convert probabilities to 1 or 2
predicted <- levels(trainfold$response)[predicted]

# We compare with the observed values and calculate error rate
observed    <- testfold$response

# Our guess on the general error rate of the model (very unprecise!)
(test_error <- sum(observed!=predicted)/length(observed)) 

```

# PCA overview 




# Predict the real unknown data

First we fit the model to all of our known data

Then we predict on the unknown data

The predictions must have the following column and the row order must be the same as the original!

* predicted (the predicted class, either "Normal" or "Tumor")

```{r}

fit <-  glm(response ~ pc2 * pc3, data = data_train, family=binomial(link='logit'))

predicted <- predict(fit, newdata = data_predict, type = "response")
predicted <- round(predicted)+1 # Convert probabilities to 1 or 2
predicted <- levels(data_train$response)[predicted]

submission <- tibble(predicted = predicted)

head(submission)

```



```{r}
# Load the necessary library
library(randomForest)

# Fit the model to the training data
fit <- randomForest(response ~ pc2 + pc3, data = trainfold)

# Predict on the test data and calculate test error
predicted_test <- predict(fit, newdata = testfold)
test_error <- mean(predicted_test != testfold$response)

# Print the test error
print(test_error)

# Fit the model to the entire training data
fit <- randomForest(response ~ pc2 + pc3, data = data_train)

# Predict on the prediction data
predicted <- predict(fit, newdata = data_predict)

# Create a tibble for the predicted values
submission <- tibble(predicted = predicted)

# Print the first few rows of the submission
head(submission)
```







# Submitting your answer

The following code will give us

* your chosen team name
* the name of the people on the team
* your estimated error rate (from train/test or CV or similar)
* your predictions

Please edit the values below .

The filename of the output will be automated minitcga_cancer_classification.TEAMNAME.rds

Please - do not use space or funny letters in your team name.

```{r}
# Load the RDS file
data <- readRDS("prediction/TCGA_cancer_classification/python_minitcga_cancer_classification.The_Brogrammers.rds")

# Format team_people
team_people <- gsub("^\\['|'\\]$", "", data$team_people[[1]][1])
team_people <- gsub("'", "", team_people)
team_people <- gsub(", ", ",", team_people)

team_error <- data$team_error[[1]][1]

# Format team_predictions
team_predictions <- gsub("^\\['|'\\]$", "", data$team_predictions[[1]])
team_predictions <- gsub("'", "", team_predictions)
team_predictions <- strsplit(team_predictions, split = ",")[[1]]
team_predictions <- trimws(team_predictions)
team_predictions <- tibble::tibble(predicted = team_predictions)

# Save all the stuff in one object
saveRDS(list(team_name, team_people, team_error, team_predictions), 
        file = paste0("minitcga_cancer_classification.", team_name, ".rds"))

```

# Checking format of all saved objects

```{r}

files   <- Sys.glob("prediction/TCGA_cancer_classification/minitcga_cancer_classification.*.rds")
results <- tibble(filename = files)

for (i in 1:nrow(results)) {
  x <- read_rds(file = as.character(results$filename[i]))
  results$team_name[i]     <- x[[1]]
  results$team_people[i]   <- paste(x[[2]], collapse=",", sep=" ")
  results$team_error[i]    <- x[[3]]
  y                        <- x[[4]]
  results$n_tumor          <- sum(y$predicted=="Tumor")
  results$n_normal         <- sum(y$predicted=="Normal")
  results$submission[i]    <- list(x[[4]])
}

rm(x,y)

results %>% select(-filename)

results$submission[[1]]

```

# Upload your rds file!
